# Plotting in `ggplot2`

## Creating high quality graphics

Once you’ve completed your spatial data analysis you’re going to need to visualise it in some really nice figures for publication and/or presentations.   

While you can create plots through various ways, including base `R`, the most popular method of producing fancy figures is with the `ggplot2` package. You can plot **raster** and **vector** spatial data with `ggplot2`. Vector data represent geographical phenomena by *points*, *lines* and *polygons*. Raster is another representation of spatial data that consist of pixels.   

First things first, if you haven’t done so yet, we need to install the `ggplot2` package. You can install `tidyverse` package that is a collection of packages including `ggplot2`.

```{r eval=FALSE}
install.packages("tidyverse")
```

## Grammar of Graphics

`ggplot2` is built on the grammar of graphics concept: the idea that any plot can be expressed from the same set of components:
  1. A **data** set
  2. A **coordinate** system
  3. A set of **geoms** - the visual representation of data points

The key to understanding `ggplot2` is thinking about a figure in layers.

To start with we need to create the base layer for our plot. This is done with the `ggplot()` function, and it is here we define our **data** and **coordinate system** components. We set our **data** component using the data argument, and then use the aesthetic function `aes()` as a second argument to define our **coordinate system**. The `aes()` function can potentially be used multiple times when constructing our figure, so it is important to know that anything defined inside the base `ggplot()` function is a *global option* inherited by all subsequent layers.

```{r}
# loading the libraries
library(sf)
library(raster)
library(tidyverse) # this includes ggplot
library(viridis) # for colour scheme
```

## Plotting spatial data in ggplot

```{r}
# loading world map
library(rworldmap)

wmap <- rworldmap::getMap(resolution = "low")  %>%
  st_as_sf()
```

### Plotting vector data

Plotting vector data with `sf` is very easy! You only need to provide the data with a *sf object* and use `geom_sf`. No need to specify `x` and `y` coordinates.

```{r}
ggplot(data = wmap) +
  geom_sf()
```

Now, the colour and background can be changed.

```{r}
ggplot(data = wmap) +
  geom_sf(aes(fill = NAME)) +
  theme_minimal() +
  guides(fill = FALSE) # don't show legend
```

### Different spatial reference in plotting

Use `coord_sf` to specify the projection that map is going to be plotted with.

```{r}
# see the coordinate system string
st_crs(3035)

ggplot(data = wmap) +
  geom_sf() +
  theme_minimal() +
  coord_sf(crs = 3035)
```

You can aslo change the colour of background.

```{r}
# read Australia polygon shapefile
au <- st_read("data/AU_states.shp", 
              quiet = TRUE)

ggplot(data = au) +
  geom_sf(size = 0.3, 
          fill = "white") + # change the line size and polygon colours
  theme_minimal() +
  coord_sf(crs = 3112) + 
  theme(panel.background = element_rect(fill = "steelblue2"))
```

### Plotting raster data

Raster can be large files containing millions of pixels. When you plot this data, the computer can not visualise all the details in the data. Plotting with full dataset could also be time-consuming. To plot rasters, it's better to get a large sample of the data, then plot it. If the sample is large enough, the plotting quality won't decrease.

```{r}
# read Australia polygon shapefile

au <- st_read("data/AU_states.shp", 
              quiet = TRUE)

# reading raster data and clip and mask it for Australia

au_temp <- raster("data/wc_tavg/wc2.0_2.5m_tavg_01.tif") %>% 
  crop(au)

# regular sample of 500k cells

sam <- sampleRegular(au_temp, 
                     size = 5e5, 
                     asRaster = TRUE) %>% 
  as.data.frame(xy = TRUE, 
                na.rm = TRUE) %>% 
  setNames(c("x", 
             "y", 
             "layer"))

head(sam)

ggplot(data = sam, 
       aes(x = x, 
           y = y, 
           fill = layer)) +
  geom_raster()
```

You can change the colour and other parametrs of the plot to make it better. It is important to use `coord_sf` or `coord_fixed` to keep the dimention of the plot fixed.

```{r}
ggplot(data = sam, 
       aes(x = x, 
           y = y, 
           fill = layer)) +
  geom_raster() +
  scale_fill_viridis(option = "A") +
  labs(x = "Longitude", 
       y = "Latitude", 
       fill = "Temperature") +
  coord_sf(crs = 4326) +
  theme_bw()
```
