---
title: "Introduction to Spatial analysis with R"
subtitle: "Raster data, Interactions and Plotting"
author: "Jutta Beher, Roozbeh Valavi and David Wilkinson"
date: "15/07/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
eval = FALSE
message = FALSE
warning = FALSE
```

# Rasters and raster analysis

In this part, you learn what a raster is, and how to read, plot and manipulate such data


First install the required packages


```{r, message = FALSE, warning = FALSE}
library(rgdal)
library(raster)
library(tidyverse)
library(sf)
library(maps)  # for making a scale bar
library(rgeos) # for labels on polygons
```

now set the working directory to your own computer

```{r}
setwd("/Users/juttabeher/Documents/phd_melbourne/workshops/EE_Intro_Spatial_Workshop")
```


# understanding what a raster is

What is a raster? Nothing more than a matrix!
The command "raster()" turns any matrix into a raster-object that can be used for
spatial analysis
You will now create your own raster:
Create an empty matrix with the required dimensions
```{r}
r <- matrix(0, nrow=3, ncol=3)
```

Populate each row of the matrix with information (the values of the cells)
and use the "raster()" command to turn it into a raster object, which you can plot
```{r}
r[,1] <- c(1,2,3)  
r[,2] <- c(5,6,7)
r[,3] <- c(8:10)
rr <- raster(r)
r
plot(rr)
```
As you can see, every cell has a color according to its value.
You will find out later how to select specific colors.

# You can do any maths with a raster!

```{r}
plot(rr * 2)
plot(rr*100)
plot(rr + rr)
plot(rr-rr)
```

# Investigate rasters

You can also use these commands to investigate different aspects of the data
```{r}
# you can use these commands to check for specific attributes of the raster:
res(rr) # returns the cell size dimensions (an x and y value)
ncol(rr) # the number of columns
nrow(rr) # the number of rows
ncell(rr) # the number of cells (rows * columns)
dim(rr) # the dimensions of the raster (number of columns, number of rows, number of bands)
crs(rr) # the projection of the raster
xmin(rr) # the x and y minimum and maximum coordinates, if there are any
ymin(rr) 
xmax(rr) 
ymax(rr)
range(values(rr), na.rm=TRUE) # the range of raster values, ignoring NoData
cellStats(rr, mean) # the mean, min, max, sd, sum
freq(rr) # for categorical rasters, the frequency of values
```


# reading a raster into R

Now we use a raster version of the outline of Australia to extract data from the 
worldclim dataset, and create one layer for mean annual temperature, and mean annual
rainfall 

The raster command is used to read in single data 
```{r}
au <- raster("data/australia.tif")

plot(au) 
```


# Reading a list of rasters into R

However, often you have to read in a lot of files at the same time, for example when working with timeseries such as climate data that often come in files for single months.
instead of reading in every single file individually, you can create "stacks".
This is particulalry useful if you intend to combine them, for example by
calculating the annual mean from 12 monthly means.
To do so, you can create a list from all files in a directory to read in at once, and then create a "stack" of raster, like a thick sandwich.


```{r}
rasterlist_t <- list.files("data/wc_tavg", pattern = ".tif$", full.names = TRUE)
t_stack <- stack(rasterlist_t)
plot(t_stack) # you should see that there are 12 rasters
```

Now do the same for the rainfall data!

```{r}
#don't show code 
rasterlist_p <- list.files("data/wc_prec", pattern = ".tif$", full.names = TRUE)
p_stack <- stack(rasterlist_p)
```

# Challenge One: Operations on rasterstacks

Now you can use any command you want ONCE on this big sandwich of rasters 
instead of having to use it 12 times on the single files.
The %>% (pipe command) works like that: after the first commmand is executed, 
the output gets used for the second command.
That can be tranlated like this: Clipping a raster is a two step process. First, the raster gets cropped to the rectangular extent of the data file used for the clipping (crop function), and then all the cells that have no values in the data file that is used for the clipping are set to NoData (mask function). you can check out if you can see the difference if you don't use the mask command


```{r}
t_stack_crop <- crop(t_stack, au) 
t_stack_crop_mask <- mask(t_stack_crop, au)
plot(t_stack_crop_mask)
```

You can speed things up with a pipe:

```{r}
t_stack_au <- crop(t_stack, au) %>% 
  mask(au)
p_stack_au <- crop(p_stack, au) %>% 
  mask(au)
plot(p_stack_au)
```


Now you can caluclate the mean from your cropped stack
```{r}
# don't show code)
pmean <- mean(p_stack_au)
tmean <- mean(t_stack_au)
```

You can use the plot() command to check how your output looks
or get a summary by exectuting the name of your new raster

```{r}
# don't show code
plot(tmean)
plot(pmean)

```

Well done, you have mastered single rasters, raster stacks and creating new information from your data!

# Writing new rasters to file
You can write your new data to file if you want.

```{r}
writeRaster(tmean, "tmean.tif", datatype = "GTiff", overwrite = TRUE)
```

# Challenge 2: creating categories from summarizing and combining data

Now you want to make categories for the temperature and the rainfall to get a better overview of distinct combinations. For this exercise, create 4 categories for temperature: cold, cool,warm, hot
and 2 for rainfall: dry and wet
Tip: use the cellStats(x, stats) command to find out the quantiles as boundaries
for temperature and min,mean,max as boundaries for the rainfall
How you do it:
Maybe you can remember how we created our first raster in the beginning - you 
create such a raster to use for reclassification. It has to show the start and endpoints for your categories (that you found out with CellStats()), as well as the numbers you want to use as category-names

```{r}
tcat <- matrix(0, nrow=4, ncol=3)
# populate each row of the matrix with reclassification information for temperature
tcat[,1] <- c(4,19,22,25) # fill in here the start of the categories 
tcat[,2] <- c(19,22,25,30) # fill in here the endpoint of the categories
tcat[,3] <- c(1:4) # fill in here the number for the categories (1-4)


pcat <- matrix(0, nrow=2, ncol=3)
# populate each row of the matrix with reclassification information for precipitation
pcat[,1] <- c(6,40) 
pcat[,2] <- c(40,380)
pcat[,3] <- c(1:2)
```


# Reclassication
Now you can use the matrix for reclassification.
Reclassification means, that a raster will get overwritten with new values in all cells, based on bins of values. You just created these bins with your matrix. For example, all cells between 4 and 19 degrees will get assigned to category 1 in the new raster.
```{r}
tcat_au <- reclassify(tmean, tcat, right=FALSE, overwrite = TRUE)
# do the same for precipitation
# don't show code
pcat_au <- reclassify(pmean, pcat, right=FALSE, overwrite = TRUE)

# check out what you created
plot(tcat_au)
plot(pcat_au)
```

Now you can combine the 2 rasters to get 8 categories for combinations of temperature and rainfall...
In order to be able to tell them apart, you can multiply one raster by 10 before combining them - what would happen if you just add them up?

```{r}

climate_au <- tcat_au + pcat_au * 10

plot(climate_au)

# you can select colors for your gradients
# define the list of colour names (see Appendix for colour names)
colors <- c("white", "yellow", "lightgreen","darkgreen", "lightblue", "blue", "purple","red")
breaks <- c(11,12,13,14,21,22,23,24,25)
# plot the raster
plot(climate_au,breaks = breaks, col=colors)

```



# INTERACTION:
Now we want to find out 2 things:
1) how many koalas are in which temperature zone?
2) how are the temperature zones distributed across the states?
To find out, we have to intersect the points and polygons of the vector data with 
the raster data. This can be done with the command "extract()" from the raster package

```{r}

states <- st_read("data/AU_states.shp")
states1 <- readOGR("data/AU_states.shp") # this is an alternative to read in vector data, some other commands do not work with st objects (see mapping challenge)
as(states, "Spatial")
plot(states["STATE"])
```

# Challenge 1: which climate do koalas like?
When we extract data to points, each point will fall within one cell, so we
don't have to worry about summarizing data

```{r}
# reading csv file
koala <- read.csv("data/koala.csv")
# convert to sf
koala_sf <- st_as_sf(koala, coords = c("Longitude", "Latitude"), crs = 4326)
crs(koala_sf)  
crs(climate_au)
st_crs(koala_sf) <- crs(climate_au) # if they are the same, but spelled differently, or one is not defined
koala_climate <- raster::extract(climate_au, koala_sf, na.rm=TRUE) 

```


Check your results: Are there NA values? Why do you think there are NA values?

Which climate do koalas like?
```{r}
hist(koala_climate)
```

# Challenge 2: Extract rasters to polygons: how to summarize more complex extractions
But if we extract data to polygons, we have potentially a lot of different values 
within each state, so we have to decide what we want to do with these values
Note: if your vector data is not an sf-object, you might need to use spTransform instead when reprojecting

```{r}
crs(states)
crs(climate_au)
states_pr <- st_transform(states, crs(climate_au))
state_climate <- raster::extract(climate_au, states_pr, factors = TRUE, df=TRUE, na.rm=TRUE) # Note: extract() exists in tidyverse AND raster, so you have to specify which one you want to use
```
Now we want to sum up all cells within each state for each category

```{r}
class.counts <- as.data.frame(table(state_climate$ID,state_climate$layer))
```

How does your output look like if you don't use the "as.data.frame" command? Why do you think it is more useful as a data.frame?


# Attach the new information to your shapefile
To be able to plot this, we join our results onto our shapefile as new columns
This is slightly more advanced, so don't worry if you don't get this bit quite yet
```{r}
class.counts <- table(state_climate$ID,state_climate$layer) %>% 
  as.matrix.data.frame() %>% 
  as.data.frame() %>% 
  setNames(paste0("class_", c(11,12,13,14,21,22,23,24)))

class.counts$OBJECTID <- as.numeric(rownames(class.counts)) # 1:8 not recommended

states_pr <- left_join(states_pr, class.counts, by = "OBJECTID")

```

Now we can create maps that show the cellcount of the different climate categories in each state. You can plot single maps, or a combination

```{r}
plot(states_pr["class_14"])
plot(states_pr[19:22])
```


# Count points in polygons
We can also count how many Koalas are in each state, and attach the information
in the attribute table of the states:
```{r}
state_koala <- st_intersection(koala_sf, states_pr) 
state.counts <- as.data.frame(table(state_koala$STATENAME,state_koala$Species))
states_pr$koala_count <- state.counts$Freq

print(state.counts)

```

If you don't always work with package sf, and you still want to be really sure that things that you calculated will get filled in the right columns and rows, you can aways do a merge:

```{r}
states2 <- merge(states1, state.counts, by.x = "STATENAME", by.y = "Var1",
                             sort = TRUE)
head(states2@data)
```


# PLOTTING
You will use the basic R plotting to make a map showing climate categories, country and state outlines and points for koalas observations and then save the map to a png file

# Challenge 1: Make a simple map
```{r}
# plot to set extent
par(mar = c(2,2,2,2))
plot(climate_au, legend=F, col=colors, main="Koala occurrence in different states and climates", key.pos = NULL)
# plot with colors for levels of factor (filled / outline)

# add the state outlines
plot(states2, add=TRUE, col=NA, border="red")# play with col and border!
polygonsLabel(states2, labels = states2$Freq, method = "centroid", col = "red")
plot(koala_sf, add=TRUE, pch = 10) #,cex is a rough SHOULD WE ADD A COLUMN FOR E.G. male/female to have them add color categories?
# way to change the size of symbols or text, pch sets the symbol, feel free to play around
#map.axes(cex.axis=0.8) # here cex specifies font size # only needed if not there yet
maps::map.scale(120,-40, relwidth = 0.25, metric = TRUE,
                ratio = FALSE, cex=0.6) # exchange TRUE and FALSE to see what they

colors <- c("white", "yellow", "lightgreen","darkgreen", "lightblue", "blue", "purple","red")
leg <- c("dry cold", "dry cool", "dry warm", "dry hot",
            "wet cold", "wet cool", "wet warm", "wet hot")# c(11,12,13,14,21,22,23,24)

par(xpd=NA) # to enable to plot outside of plotting area
legend(155, y=-15, legend = leg, fill = colors)

```

